---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "John Rooney"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(lubridate)
library(here)

raw.data = "Data/Raw"
lakes.data<-read.csv(
  here(raw.data, "NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), stringsAsFactors=T)

lakes.data$sampledate <- mdy(lakes.data$sampledate)

#2
my.theme <- theme_light(base_size = 18)+
  theme(axis.text = element_text(color = "grey19"), 
        legend.position = "top",
        legend.justification = "left")
theme_set(my.theme) 
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperatures recorded during July do not change with depth across all lakes. 
Ha: At least one mean lake temperature recorded during July will change with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lakes.data.July <- lakes.data %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  filter(daynum %in% c(183:213))%>%
  drop_na()

#5
ggplot(lakes.data.July, aes(x=depth, y=temperature_C)) +
  geom_point()+
  geom_smooth(method=lm)+
  ylim(0,35)+
  labs(x="Depth in Meters", y=expression(Temperature*degree*C), 
       title="Temperature by Depth in Several Wisconsin Lakes")

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatterplot suggests that temperature is indeed affected by depth, with temperature decreasing as depth increases. The distribution of points suggests a strong negative correlation.  


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp.regression <- lm(data = lakes.data.July, temperature_C ~ depth)
summary(temp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The linear regression model shows that depth is significant at the 0.001 level, and that for every meter in depth we should expect to see a decrease in temperature of 1.95 degrees celsius. Our multiple R-squared value tells us that depth can explain 73.71% of the variability in temperature. This finding is based on 1 and 9,671 degrees of freedom. Our overall p-value for the regression is <0.001, telling us that we can reject the null hypothesis in favor of the alternate hypothesis. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp.AIC <- lm(data = lakes.data.July, temperature_C ~ depth + daynum + 
              year4)
step(temp.AIC)

#10
temp.regression2 <- lm(data = lakes.data.July, temperature_C ~ year4 + depth + daynum)
summary(temp.regression2)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables that the AIC method suggests we use includes daynum, year4, and depth. This model can explain 73.98% of the variation we see in the data. This is a small improvement of .27%. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
#anova
lake.temps.anova <- aov(data = lakes.data.July, temperature_C ~ lakename)
summary(lake.temps.anova)

#lm
lake.temps.anova2 <- lm(data = lakes.data.July, temperature_C ~ lakename)
summary(lake.temps.anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Both models, the ANOVA and linear model, show us that there is a significant difference in temperature between the lakes. In the case of the ANOVA, we can see that lakename as a variable has a p-value <0.001, telling us that lakename is a significant variable in determining temperature. The linear model shows us a bit more detail, breaking out each lake and providing estimates, standard errors, t-values, and p-values. All lakes except for Crampton Lake show as significant at the 0.001 level, telling us that they are a very significant variable when accounting for change in temperature. Crampton Lake still shows at significant, but at a 0.01 level. 

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ggplot(lakes.data.July, aes(x=depth, y=temperature_C, color=lakename)) +
  geom_point(alpha=0.5)+
  geom_smooth(method=lm)+
  ylim(0,35)+
  labs(x="Depth in Meters", y=expression(Temperature*degree*C), 
       title="Temperature by Depth in Several Wisconsin Lakes")

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(lake.temps.anova)

lake.temps.groups <- HSD.test(lake.temps.anova, "lakename", group = TRUE)
lake.temps.groups
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake seem to have the same mean temperature, statistically speaking, as Peter Lake. No lake has a mean temperature that is statistically distinct from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just looking at Peter and Paul Lakes, we could use a two-sample t-test to explore whether they have distinct mean temperatures. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
lakes.temp.July.CramptonWard <- lakes.data.July %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

lake.temps.twosample <- t.test(lakes.temp.July.CramptonWard$temperature_C ~ lakes.temp.July.CramptonWard$lakename)
lake.temps.twosample
```

>Answer: With a p-value of 0.1961, the test tells us that there is not a statistically significant difference in the average temperatures of Crampton and Ward Lakes in July. This matches both with my answer to part 16 as well as the Tukey test, which shows them as part of overlapping groups (each with a "b" designation).
